Practical Optional Types for Clojure

### 2015-07-11 formal-model.tex
add GT citations. better update figure and section 3.6

\paragraph{Expressions} Syntax is given in \figref{main:figure:termsyntax}. Expressions \e{} 
include variables \x{}, values \v{},
applications, abstractions, conditionals, and let expressions.
All binding forms introduce fresh variables.
Values include booleans \bool{}, \nil{}, class literals {\class{}}, keywords \k{},
numbers {\num{}},
constants {\const{}}, and strings \str{}. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
close value environments \openv{} over functions, which map bindings to values.

@@

\begin{figure}
  \footnotesize
$$
\begin{array}{lrll}
  \e{} &::=& \x{}
                      \alt \v{} 
                      \alt {\comb {\e{}} {\e{}}} 
                      \alt {\abs {\x{}} {\t{}} {\e{}}} &\mbox{Expressions} \\
                      &\alt& {\ifexp {\e{}} {\e{}} {\e{}}}
                      %\alt {\trdiff{\doexp {\e{}} {\e{}}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}\\
                      %\alt {\errorvalv{}}
  \v{} &::=&          \singletonmeta{}
                      \alt {\num{}}
                      \alt {\const{}}
                      \alt {\str{}}
                      \alt {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
                &\mbox{Values} \\
                \constantssyntax{}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                &\mbox{Types} \\
                      &\alt& {\Value \singletonmeta{}} 
                      \alt \trdiff{\class{}}\\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}\\
  \propenvsyntax{}\\
  \openvsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{array}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\label{main:figure:termsyntax}
\end{figure}

@@

  \end{mathpar}
  \caption{Core Subtyping rules}
  \label{main:figure:subtyping}
\end{figure}

@@

\paragraph{Typing rules}
The core typing rules are
given as \figref{main:figure:othertypingrules}. We introduce
the interesting rules with the complement number predicate
as a running example.
\begin{equation}
\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
\end{equation}
%, including
%a subsumption rule T-Subsume and rules for the false values---T-Nil and T-False---encoded 
%as impossible (\botprop{}) `then' propositions.

@@

First, in T-App the operator \e{} = \numberhuh{} is checked with T-Const, which
uses 
\constanttypeliteral{} to type constants.
\numberhuh{} is a predicate over numbers, and
\classconst{} returns its argument's class.
$$
\constanttypefigure{}
$$
The dynamic semantics for constants are given as \figref{main:figure:primitivesem}.

@@

\label{formalmodel:proofsystem}
\subsection{Proof system}

% checks obvious prop logic, but update is most interesting rule.

\begin{figure*}
  $$
\begin{array}{lllll}
\updatefigure
\end{array}
$$
\caption{Type Update}
\label{main:figure:update}
\end{figure*}

%\begin{figure}
%  $$
%\begin{array}{lllll}
%  \restrictremovefigure{}
%\end{array}
%  $$
%  \caption{Restrict and Remove}
%  \label{main:figure:restrictremove}
%\end{figure}

>>>>

\paragraph{Expressions} Syntax is given in \figref{main:figure:termsyntax}. Expressions \e{} 
include variables \x{}, values \v{},
applications, abstractions, conditionals, and let expressions.
All binding forms introduce fresh variables.
Values include booleans \bool{}, \nil{}, class literals {\class{}}, keywords \k{},
integer numbers {\nat{}},
constants {\const{}}, and strings \str{}. Lexical closures {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
close value environments \openv{} over functions, which map bindings to values.

@@

\begin{figure}
  \footnotesize
$$
\begin{array}{lrll}
  \e{} &::=& \x{}
                      \alt \v{} 
                      \alt {\comb {\e{}} {\e{}}} 
                      \alt {\abs {\x{}} {\t{}} {\e{}}} &\mbox{Expressions} \\
                      &\alt& {\ifexp {\e{}} {\e{}} {\e{}}}
                      %\alt {\trdiff{\doexp {\e{}} {\e{}}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}\\
                      %\alt {\errorvalv{}}
  \v{} &::=&          \singletonmeta{}
                      \alt {\nat{}}
                      \alt {\const{}}
                      \alt {\str{}}
                      \alt {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
                &\mbox{Values} \\
                \constantssyntax{}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                &\mbox{Types} \\
                      &\alt& {\Value \singletonmeta{}} 
                      \alt \trdiff{\class{}}\\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}\\
  \propenvsyntax{}\\
  \openvsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{array}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\label{main:figure:termsyntax}
\end{figure}

@@

  \end{mathpar}
  \caption{Core subtyping rules}
  \label{main:figure:subtyping}
\end{figure}

@@

\paragraph{Typing rules}
The core typing rules are
given as \figref{main:figure:othertypingrules}. We introduce
the interesting rules with the complement natural number predicate
as a running example.
\begin{equation}
\abs{\makelocal{d}}{\Top}{\ifexp{\appexp{\numberhuh{}}{\makelocal{d}}}{\false{}}{\true{}}}
\end{equation}
%, including
%a subsumption rule T-Subsume and rules for the false values---T-Nil and T-False---encoded 
%as impossible (\botprop{}) `then' propositions.

@@

First, in T-App the operator \e{} = \numberhuh{} is checked with T-Const, which
uses 
\constanttypeliteral{} (\figref{main:figure:constanttyping})
to type constants.
\numberhuh{} is a predicate over the natural numbers, and
\classconst{} returns its argument's class.

\begin{figure}
  $$
\constanttypefigure{}
  $$
  \caption{Constant typing}
  \label{main:figure:constanttyping}
\end{figure}

The dynamic semantics for constants are given as \figref{main:figure:primitivesem}.

@@

\begin{figure*}
  $$
\begin{array}{llll}
\updatefigure{}
\end{array}
$$
$$
\begin{array}{lr}
  \begin{array}{llll}
    \restrictfigure{}
  \end{array}
  \begin{array}{llll}
    \removefigure{}
  \end{array}
\end{array}
$$
\caption{Type update (the metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables), 
  \notsubtypein{}{\Nil{}}{\nottype{\t{}}} when \issubtypein{}{\Nil{}}{\t{}})}
\label{main:figure:update}
\end{figure*}

%\begin{figure}
%  $$
%\begin{array}{llll}
%  \restrictremovefigure{}
%\end{array}
%  $$
%  \caption{Restrict and Remove}
%  \label{main:figure:restrictremove}
%\end{figure}

\subsection{Proof system}
\label{formalmodel:proofsystem}

\begin{figure}
  $$
\LUpdate{}
$$
%(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
  \caption{Proof system update rule (full rules \figref{appendix:figure:proofsystem})}
  \label{main:figure:lupdate}
\end{figure}

The proof system for occurrence typing uses standard propositional logic,
except for where nested information is combined. This is
handled by the L-Update rule (\figref{main:figure:lupdate})---it says
under \propenv{}, if object \path{\pathelemp{}}{\x{}} is of type \t{}, and 
some extension
\path{\pathelem{}}{\path{\pathelemp{}}{\x{}}}
is of possibly-negative type \propisnotmeta{}, then
{\update{\t{}}{\propisnotmeta{}}{\pathelem{}}}
is the new type of \path{\pathelemp{}}{\x{}} under \propenv{}.

Recall \egref{example:desserts-on-meal}.
%, resuming from
%\secref{sec:coretypesystem}. 
Solving
$
{ \inpropenv 
  {{\isprop{\Order}{\makelocal{o}}},
    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}}}
  {\isprop {\t{}} {\makelocal{o}}}}
$
uses L-Update, where \pathelem{} = {\emptypath{}} and \pathelemp{} = [{\keype{\makekw{Meal}}}].
$$
\inpropenv{\propenv{}}{\isprop{\update{\Order}{\Value{\makekw{combo}}}{[{\keype{\makekw{Meal}}}]}}{\makelocal{o}}}
$$
Since {\Order} is a union of HMaps, we structurally recur on the first case of \updateliteral{}
(\figref{main:figure:update}),
which preserves \pathelem{}.
Each initial recursion hits the first HMap case, since there is some \t{} such that
{\inmandatory{\k{}}{\t{}}{\mandatory{}}} and 
\completenessmeta{} accepts partial maps \partial{}.

To demonstrate,
\makekw{lunch} meals hit the first HMap case and
update to {\HMapp {\extendmandatoryset {\mandatory{}}{\Valkw{Meal}}{\sp{}}} {\emptyabsent{}}}
where \sp{} = {\update{\Valkw{lunch}}{\Valkw{combo}}{\emptypath{}}}
and \mandatory{} = \mandatoryset{\mandatoryentry{\Valkw{Meal}}{\Valkw{lunch}},{\mandatoryentry{\Valkw{desserts}}{\Number{}}}}.
\sp{} updates to \Bot via the penultimate \updateliteral{} case,
because \restrict{\Value{\makekw{lunch}}}{\Value{\makekw{combo}}} = \Bot
by the first \restrictliteral{} case.
The same happens to \makekw{dinner} meals,
leaving just the \makekw{combo} HMap.

In \egref{example:desserts-on-class},
$
\inpropenv{\propenv{}}{\isprop{\update{\Order}{\Long}{[{\classpe{}}, {\keype{\makekw{desserts}}}]}}{\makelocal{o}}}
$
updates the argument in the {\Long} method.
This recurs and extra time for each meal to handle the {\classpe{}}
path element.
%$
%{\update{\Number}{\Long}{[{\classpe{}}]}}}
%$
%
%$
%{\update{\Int}{\Long}{[{\classpe{}}]}}}
%$
%
%

### 2015-07-11 overview.tex
add GT citations. better update figure and section 3.6

The following Typed Clojure program constructs a new \clj{File}
instance and calls the \clj{getParent} method on the result.

\begin{exmp}
\begin{minted}{clj}
  ; :- (U nil Str)
  (.getParent (new File "a/b"))  
  ;=> "a"
\end{minted}
\label{example:getparent-direct-constructor}
\end{exmp}

@@

\begin{minted}{clojure}
(ann desserts' [Order -> Int])
(defmulti desserts' 
          (fn [o :- Order] (class (:desserts o))))
(defmethod desserts' Long [{d :desserts :as o}] 
  ; o :- (U '{:Meal ':lunch :desserts Int}
  ;         '{:Meal ':dinner :desserts Int})
  d)
(defmethod desserts' nil [{m1 :meal1 m2 :meal2 :as o}]
  ; o :- '{:Meal ':combo :meal1 Order :meal2 Order}
  (+ (desserts' m1) (desserts' m2)))
\end{minted}
%
%(desserts' {:Meal :combo 
%            :meal1 {:Meal :lunch :desserts 1}
%            :meal2 {:Meal :dinner :desserts 2}})
%;=> 3

In the \clj{Long} method, Typed Clojure learns that
its argument is at least of type \clj{'{:desserts Long}}---since
\begin{minted}{clojure}
(isa? (class (:desserts o)) Long)
\end{minted}
must be true.
%
Knowing \clj{o} is also an
\clj{Order}, we can eliminate the possibility of
a \clj{:combo} meal, and deduce the type:
\begin{minted}{clojure}
(U '{:Meal ':lunch :desserts Int}
   '{:Meal ':dinner :desserts Int})
\end{minted}
Looking up \clj{:desserts} on this type returns an \clj{Int}.

In the \clj{nil} case, we know
\begin{minted}{clojure}
(isa? (class (:desserts o)) nil)
\end{minted}
is true---which implies \clj{(class (:desserts o))} is \clj{nil}.

Since lookups on missing keys return \clj{nil}, either
\begin{itemize}
  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, or
  \item \clj{o} is missing a \clj{:desserts} entry.
\end{itemize}
Equivalently, we learn \clj{o} is at least of type
\begin{minted}{clojure}
(U '{:desserts nil}
   ; :absent-keys, a set of known absent entries
   (HMap :absent-keys #{:desserts}))
\end{minted}
Since \clj{o} is originally an \clj{Order}, this simplifies to a \clj{:combo} meal, 
\begin{minted}{clojure}
'{:Meal ':combo :meal1 Order :meal2 Order}
\end{minted}
thus allowing both recursive calls to type check.

>>>>

The following Typed Clojure program constructs a new \clj{File}
instance and calls the \clj{getParent} method on the result,
returning a string \clj{"a"}, inferred as nullable.

\begin{exmp}
\begin{minted}{clj}
(.getParent (new File "a/b"))  
; :- (U nil Str)
;=> "a"
\end{minted}
\label{example:getparent-direct-constructor}
\end{exmp}

@@

In the first method, the dispatch value is the concrete class \clj{Long}, a subtype
of \clj{Int}, and the second \clj{nil}, the sentinel value for a failed map lookup.
In practice, \clj{:lunch} and \clj{:dinner} meals will dispatch to the \clj{Long}
method, but Typed Clojure infers a slightly more general type due to the definition
of \clj{:combo} meals.

\begin{exmp}
\begin{minted}{clojure}
(ann desserts' [Order -> Int])
(defmulti desserts' 
          (fn [o :- Order] (class (:desserts o))))
(defmethod desserts' Long [{d :desserts :as o}] 
  ; o :- (U '{:Meal ':lunch :desserts Int}
  ;         '{:Meal ':dinner :desserts Int}
  ;         '{:Meal ':combo :desserts Int
  ;           :meal1 Order :meal2 Order})
  d)
(defmethod desserts' nil [{m1 :meal1 m2 :meal2 :as o}]
  ; o :- '{:Meal ':combo :meal1 Order :meal2 Order}
  (+ (desserts' m1) (desserts' m2)))
\end{minted}
\label{example:desserts-on-class}
\end{exmp}
%
%(desserts' {:Meal :combo 
%            :meal1 {:Meal :lunch :desserts 1}
%            :meal2 {:Meal :dinner :desserts 2}})
%;=> 3

In the \clj{Long} method, Typed Clojure learns that
its argument is at least of type \clj{'{:desserts Long}}---since
\begin{minted}{clojure}
(isa? (class (:desserts o)) Long)
\end{minted}
must be true.
%
%Knowing \clj{o} is also an
%\clj{Order},
In this method, we deduce
the entry \clj{:desserts}
\emph{must} be a present and mapped to an \clj{Int},
even in a \clj{:combo} meal
which does not state \clj{:desserts}
as present or absent.

In the \clj{nil} method,
\clj{(isa? (class (:desserts o)) nil)}
must be true---which implies \clj{(class (:desserts o))} is \clj{nil}.
%
Since lookups on missing keys return \clj{nil}, either
\begin{itemize}
  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, like \clj{{:desserts nil}}, or
  \item \clj{o} is missing a \clj{:desserts} entry, like \clj{{}}.
\end{itemize}
Equivalently, we learn \clj{o} is at least of type
\begin{minted}{clojure}
(U '{:desserts nil}
   ; :absent-keys, a set of known absent entries
   (HMap :absent-keys #{:desserts}))
\end{minted}
This eliminates non-\clj{:combo} meals
since their \clj{'{:desserts Int}} type does not agree
with this new information (because \clj{:desserts}
is neither mapped to \clj{nil} or absent).

%simplifies to a \clj{:combo} meal, 
%\begin{minted}{clojure}
%'{:Meal ':combo :meal1 Order :meal2 Order}
%\end{minted}
%thus allowing both recursive calls to type check.
